The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-10-13 poster/main.tex

    % --- Column 7: All Institutions ---
    \makecell{
    \Large $^{1}$Moscow Institute of Physics and Technology\\
    \Large $^{2}$Skolkovo Institute of Science and Technology\\
    \Large $^{3}$HSE\; \Large $^{4}$AI4Science\; \Large $^{5}$AIRI
    }
  \end{tabular*}%
  \hspace*{\fill}%
}

@@

Training large language models requires optimizers more scalable and efficient than Adam/AdamW. The recent Muon~[2] optimizer shows great promise, sometimes doubling AdamW's efficiency, due to its unique design:
\begin{itemize}
  \item \textbf{\color{HazySummerEve}Matrix-Aware Design:} It is constructed specifically for weight matrices, not generic vectors.
  \item \textbf{\color{HazySummerEve}Principled Update Rule:} Muon's update is not merely heuristic. It is derived as the solution to a linear minimization problem constrained by the operator (spectral) norm, giving it a solid mathematical foundation.
\end{itemize}

Muon's success has inspired variants like Dion~[1] algorithm and frameworks of Scion and Gluon~[4]. We build on this work by asking a fundamental question:
\begin{attention}
In deriving Muon's update step, why constrain by the spectral norm? How would alternative norms affect performance and computational cost?
\end{attention}
To investigate this, we first analyze the link between Muon-like algorithms and their defining norms. We then propose \textbf{\color{HazySummerEve}F-Fanions}, a novel family of optimizers derived from exotic, mixed norms. Finally, we present an empirical comparison between Muon and F-Fanions across various numerical experiments.

@@

\centering
% Using 'tabular' now for better width control based on content
\begin{tabular}{clcc}
\toprule
\textbf{Case} & \textbf{Method} & \textbf{Constraining Norm} & \textbf{Update Formula} \\
\midrule
\multirow{3}{*}{\rotatebox[origin=c]{90}{\parbox{1.5cm}{\centering Vec.}}}
& Normalized SGD    & $\ell_2$          & $-\eta g / \norm{g}_2$ \\
& SignSGD           & $\ell_\infty$     & $-\eta\sign(g)$ \\
& Coordinate Descent    & $\ell_1$          & $-\eta e_{\mathrm{argmax}|g_i|}$ \\
\midrule
\multirow{3}{*}{\rotatebox[origin=c]{90}{\parbox{1.5cm}{\centering Mat.}}}
& Normalized SGD    & $\normf{\cdot}$   & $-\eta\mG^t / \normf{\mG^t}$ \\
& Muon     & $\norms{\cdot}$   & $-\eta\mU \mV^\top$ \\
& Dion (no feedback)   & $\norm{\cdot}_{KF-k}^\dag$ & $-\eta\mU_k \mV_k^\top$ \\
\bottomrule
\end{tabular}

@@

This defines the \textbf{\color{HazySummerEve}F-Fanions}, a new family of optimizers parameterized by $\alpha \in [0, 1]$ and $k$, with the update rule:
\begin{equation*}
    \Delta \mX^t = \eta \alpha \mU_k \mV_k^\top + \eta(1 - \alpha) \frac{\mG^t}{\normf{\mG^t}}\,.
\end{equation*}

This framework recovers existing algorithms as special cases:
\begin{itemize}
    \item \textbf{\color{HazySummerEve}Normalized SGD:} $\alpha = 0$.
    \item \textbf{\color{HazySummerEve}Dion} (no feedback): $\alpha = 1$, $k < \min\{m, n\}$.
    \item \textbf{\color{HazySummerEve}Muon:} $\alpha = 1$, $k = \min\{m, n\}$.
\end{itemize}
We also identify and analyze \textbf{\color{HazySummerEve}F-Neon}, a computationally efficient member of this family where $k=1$.

@@

We benchmarked our method against Muon on the \texttt{modded\_nanogpt\_2024} speedrun, a time-constrained training challenge. Below are the results after 1750 steps,

@@

    Another Interesting observation here is that tuned F-Muon has much larger trust region then Muon, and thus makes larger step sizes (figure to the right).
  \end{column}
  \begin{column}{0.49\textwidth}
    \includegraphics[width=\linewidth]{fstardual_cifar.pdf}
    \centering
  \end{column}
\end{columns}

@@

\begin{itemize}
    \item We introduced the \textbf{\color{HazySummerEve}F-Fanion framework}, unifying LMO-based optimizers like Muon and Dion via mixed norms.

@@

\begin{itemize}
   \large
    \setlength{\itemsep}{0pt} % Remove spacing between items
    \item[{[1]}] Ahn, K. \& Xu, B. *Dion*. arXiv:2504.05295, 2025.
    \item[{[2]}] Bernstein, J. *Deriving muon*. 2025. URL: jeremybernste.in/writing/deriving-muon
    \item[{[3]}] Bernstein, J. \& Newhouse, L. *Old optimizer, new norm*. arXiv:2409.20325, 2024.
    \item[{[4]}] Riabinin, A. et al. *Gluon: Making muon \& scion great again!* arXiv:2505.13416, 2025.
\end{itemize}

>>>>

    % --- Column 7: All Institutions ---
    \makecell{
    \Large $^{1}$MIPT\; \Large $^{2}$Skoltech\\
    \Large $^{3}$HSE\; \Large $^{4}$AI4Science\\
    \Large $^{5}$AIRI
    }
  \end{tabular*}%
  \hspace*{\fill}%
}

@@

Training large language models requires optimizers more scalable and efficient than Adam/AdamW. The recent Muon~[2] optimizer shows great promise, sometimes doubling AdamW's efficiency, due to its unique design:
\begin{itemize}
  \item \textbf{\color{HazySummerEve}Matrix-Aware Design:} It is constructed specifically for weight matrices rather than generic vectors.
  \item \textbf{\color{HazySummerEve}Principled Update Rule:} Muon's update is not merely heuristic. It is derived as the solution to a linear minimization problem constrained by the operator (spectral) norm, giving it a solid mathematical foundation.
\end{itemize}

Muon's success has inspired variants like Dion~[1] algorithm and frameworks of Scion and Gluon~[4]. We build on this work by asking a fundamental question:
\begin{attention}
In deriving Muon's update step, why constrain by the spectral norm? How would alternative norms affect performance and computational cost?
\end{attention}
To investigate this, we first analyze the link between Muon-like algorithms and their defining norms. We then propose \textbf{\color{HazySummerEve}F-Fanions}, a novel family of optimizers derived from exotic, mixed norms. Finally, we present an empirical comparison between Muon and F-Fanions.

@@

{
\centering
% Using 'tabular' now for better width control based on content
\begin{tabular}{clcc}
\toprule
\textbf{Case} & \textbf{Method} & \textbf{Norm} & \textbf{Update Formula} \\
\midrule
\multirow{3}{*}{\rotatebox[origin=c]{90}{\parbox{1.5cm}{\centering Vec.}}}
& Normalized SGD    & $\ell_2$          & $-\eta g^t / \norm{g^t}_2$ \\
& SignSGD           & $\ell_\infty$     & $-\eta\sign(g^t)$ \\
& Gauss-Southwell CD    & $\ell_1$      & $-\eta \sum_{i \in \argmax|g_i^t|} \sign(g_i^t)e_i$ \\
\midrule
\multirow{3}{*}{\rotatebox[origin=c]{90}{\parbox{1.5cm}{\centering Mat.}}}
& Normalized SGD    & $\normf{\cdot}$   & $-\eta\mG^t / \normf{\mG^t}$ \\
& Muon     & $\norms{\cdot}$   & $-\eta\mU \mV^\top$ \\
& Dion (without EF)   & $\norm{\cdot}_{KF-k}^\dag$ & $-\eta\mU_k \mV_k^\top$ \\
\bottomrule
\end{tabular}\eqspace
}
Here $\norm{\cdot}_{KF-k}$ denotes Ky Fan $k$ norm, $\norm{\cdot}^\dag$ is the dual norm.

@@

This defines the \textbf{\color{HazySummerEve}F-Fanions}, a new family of optimizers parameterized by $\alpha \in [0, 1]$ and $k$, with the update rule:
\begin{equation*}
    \Delta \mX^t = -\eta \alpha \mU_k \mV_k^\top - \eta(1 - \alpha) \frac{\mG^t}{\normf{\mG^t}}\,.
\end{equation*}

This framework recovers existing algorithms as special cases:
\begin{itemize}
    \item \textbf{\color{HazySummerEve}Normalized SGD:} $\alpha = 0$.
    \item \textbf{\color{HazySummerEve}Dion} (without error feedback): $\alpha = 1$, $k < \min\{m, n\}$.
    \item \textbf{\color{HazySummerEve}Muon:} $\alpha = 1$, $k = \min\{m, n\}$.
\end{itemize}
We also identify and analyze \textbf{\color{HazySummerEve}F-Neon}, a computationally efficient member of this family where $k=1$.

@@

We benchmarked our method against Muon on the \texttt{modded nanogpt 2024} speedrun, a pretraining of a real LLM. Below are the results after 1750 steps,

@@

    Another Interesting observation here is that tuned F-Muon has much larger trust region than Muon, and thus makes larger step sizes (figure to the right).
  \end{column}
  \begin{column}{0.49\textwidth}
    \includegraphics[width=\linewidth]{fstardual_cifar.pdf}
    \centering
  \end{column}
\end{columns}

@@

\begin{itemize}
    \item We introduced the \textbf{\color{HazySummerEve}F-Fanion framework}, unifying LMO-based optimizers like Muon and Normalized SGD via mixed norms.

@@

\begin{itemize}
   \large
    \setlength{\itemsep}{0pt} % Remove spacing between items
    \item[{[1]}] Ahn, K. \& Xu, B. \textit{Dion: A Communication-Efficient Optimizer for Large Models}. arXiv:2504.05295, 2025.
    \item[{[2]}] Bernstein, J. \textit{Deriving Muon}. 2025. URL: jeremybernste.in/writing/deriving-muon
    \item[{[3]}] Bernstein, J. \& Newhouse, L. \textit{Old Optimizer, New Norm: An Anthology}. arXiv:2409.20325, 2024.
    \item[{[4]}] Riabinin, A. et al. \textit{Gluon: Making Muon \& Scion Great Again!} arXiv:2505.13416, 2025.
\end{itemize}
